---
title: 'Analysis'
---

```{r packages}
library(tidyverse)
library(rvest)
library(httr)
library(jsonlite)
library(patchwork)

knitr::opts_chunk$set(
fig.width = 6,
fig.asp = .6,
out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
ggplot2.continuous.colour = "viridis",
ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

### Exploratory Data Analysis

## Import Data

```{r, message = FALSE}
total_state_revenue = read_csv("data/total_state_revenue.csv")
model_clean_2024 = read_csv("data/model_clean_2024.csv")
```

## Gambling visuals

# Plot gambling revenue by year

```{r Plotting gambling revenue by Year}
# Aggregate total revenue by year
yearly_revenue_summary = total_state_revenue |> 
  group_by(Year) |> 
  summarize(Total_Revenue = sum(Revenue, na.rm = TRUE))

yearly_revenue_summary |> 
  ggplot(aes(x = Year, y = Total_Revenue)) +
  geom_col(aes(fill = as.factor(Year)), show.legend = FALSE) +
  geom_text(aes(label = scales::dollar(Total_Revenue, scale = 1e-9, suffix = "B")), 
            vjust = -0.5) + # Add labels in Billions
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_x_continuous(breaks = 2018:2024) +
  labs(
    title = "Total US Gambling Revenue (2018-2024)",
    subtitle = "Aggregated revenue across all states",
    x = "Year",
    y = "Total Revenue"
  )
```
 
# Plotting gambling revenue by state over time

```{r Plotting gambling revenue by State Over time}
# Identify top 5 states by 2024 revenue to keep things looking clean
top_states = total_state_revenue |> 
  filter(Year == 2019) |> 
  slice_max(Revenue, n = 5) |> 
  pull(State)

state_growth_plot_data = total_state_revenue |> 
  mutate(
    State_Group = ifelse(State %in% top_states, State, "All Other")
  ) |> 
  group_by(Year, State_Group) |> 
  summarize(Revenue = sum(Revenue, na.rm = TRUE), .groups = "drop") |> 
  mutate(State_Group = fct_reorder(State_Group, Revenue, .desc = TRUE))


ggplot(state_growth_plot_data, aes(x = Year, y = Revenue, fill = State_Group)) +
  geom_col() +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_x_continuous(breaks = 2018:2024) +
  labs(
    title = "US Gambling Revenue Growth by State",
    x = "Year",
    y = "Revenue",
    fill = "State"
  )
```

# Plotting revenue rank: top 5 gambling markets by year since 2018

```{r plotting revenue rank: top 5 gambling markets by year since 2018}
# Calculate Ranks for every state, every year
ranked_data = total_state_revenue |> 
  group_by(Year) |> 
  mutate(Rank = min_rank(desc(Revenue))) |> 
  ungroup()

# Identify "States of Interest"
# We want to track any state that was Top 5 at any point
focus_states = ranked_data |> 
  filter(Rank <= 5) |>  # can change if we want less point to start and end points: filter((Year == 2018 & Rank <= 5) | (Year == 2024 & Rank <= 5)) |>
  pull(State) |> 
  unique()

# Filter the dataset to just those states
plot_data = ranked_data |> 
  filter(State %in% focus_states)

plot_data |> 
  ggplot(aes(x = Year, y = Rank, color = State, group = State)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) + 
  scale_y_reverse(breaks = 1:max(plot_data$Rank, na.rm = TRUE)) + 
  scale_x_continuous(breaks = 2018:2024) +
  geom_text(data = plot_data |> filter(Year == 2018), 
            aes(label = paste(Rank, State)), x = 2017.8, hjust = 1, size = 3) +
  geom_text(data = plot_data |> filter(Year == 2024), 
            aes(label = paste(Rank, State)), x = 2024.2, hjust = 0, size = 3) +
  labs(
    title = "Shift in Top Gambling Markets (2018 vs 2024)",
    subtitle = "Tracking the rank of states that were in the top 5 in either 2018 or 2024",
    y = "Revenue Rank",
    x = "Year"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",        
    panel.grid.minor = element_blank()
  ) +
  coord_cartesian(xlim = c(2017, 2025))
```

## Demographic visuals

# Plot of Total Population in 2024 by State

```{r}
population_plot = 
  model_clean_2024 |> 
    filter(State != "Washington D.C.") |> #drop DC bc all demographic data is NA
    mutate(
      State = as.factor(State),
      State = fct_reorder(State, total_pop, .desc = TRUE)
    ) |> 
    ggplot(aes(x = State, y = total_pop)) +
    geom_col(aes(fill = State)) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
      legend.position = "none") +
    labs(
        y = "Total Population",
        title = "Total Population by State in 2024"
      )
```

# Plot of Gambling Revenue in 2024 by State

```{r}
revenue_plot = 
  model_clean_2024 |> 
    filter(State != "Washington D.C.") |> #drop DC bc all demographic data is NA
    mutate(
      State = as.factor(State),
      State = fct_reorder(State, total_pop, .desc = TRUE)
    ) |> 
    ggplot(aes(x = State, y = Revenue)) +
    geom_col(aes(fill = State)) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
      legend.position = "none") +
    labs(
        x = "State (in order of largest to smallest total population)",
        y = "Revenue",
        title = "Gambling Revenue by State in 2024"
      )
```

# Patchwork Plot of Total Population and Gambling Revenue in 2024 by State

```{r, fig.width = 10}
combined_revenue_population = (population_plot + revenue_plot)

combined_revenue_population
```

# Racial Makeup of States in 2024

```{r}
model_clean_2024 |> 
  filter(State != "Washington D.C.") |> #drop DC bc all demographic data is NA
  select(State, Revenue, pct_white, pct_black, pct_asian, pct_hispanic) |> 
  pivot_longer(
    cols = pct_white:pct_hispanic,
    names_to = "race",
    values_to = "percentage"
  ) |> 
  mutate(
      State = as.factor(State),
      State = fct_reorder(State, Revenue, .desc = TRUE),
      race = factor(race,
                  levels = c("pct_white", "pct_black", "pct_asian", "pct_hispanic"),
                  labels = c("% White", "% Black", "% Asian", "% Hispanic"))
  ) |> 
  ggplot(aes(x = State, y = percentage, color = race, group = race)) +
  geom_point() +
  geom_line() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1)) +
    labs(
        x = "State (in order of largest to smallest gambling revenue)",
        y = "Percentage of Population",
        title = "Racial Makeup by State in 2024",
        fill = "Race"
    ) 
```

# Poverty % and Unemployment Rate by State in 2024

```{r}
model_clean_2024 |> 
  filter(State != "Washington D.C.") |> #drop DC bc all demographic data is NA
  select(State, Revenue, unemployment_rate, pct_poverty) |> 
  pivot_longer(
    cols = c("pct_poverty", "unemployment_rate"),
    names_to = "measure",
    values_to = "percentage"
  ) |> 
  mutate(
      State = as.factor(State),
      State = fct_reorder(State, Revenue, .desc = TRUE),
      measure = factor(measure,
                  levels = c("pct_poverty", "unemployment_rate"),
                  labels = c("% in Poverty", "% Unemployed"))
  ) |> 
  ggplot(aes(x = State, y = percentage, color = measure, group = measure)) +
  geom_point() +
  geom_line() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1)) +
    labs(
        x = "State (in order of largest to smallest gambling revenue)",
        y = "Percentage of Population",
        title = "% Poverty and Unemployment Rate by State in 2024",
        fill = "Measure"
    ) 
```

