---
title: "gambling_rev"
output:
  github_document:
    df_print: default
  html_document:
    df_print: paged
always_allow_html: true
date: "2025-11-30"
---
```{r packages}
library(tidyverse)
library(rvest)
library(httr)
library(jsonlite)
library(glmnet)
library(dplyr)
library(webshot2)

```
## Data scraping gambling revenue and growth
```{r scraping}
url = "https://rg.org/statistics/us"
gambling_html = read_html(url)
```

```{r}
gambling_html |>
  html_table()
```
### pull in 2024 revenue by state table 
```{r state revenue}
state_revenue_24 = 
  gambling_html |> 
  html_table() |> 
  first() 
```
### pull in 2025 revenue growth by month
```{r monthly growth}
revenue_growth_25= 
  gambling_html |> 
  html_table() |> 
  purrr::pluck(2)
```

The website that was scrapped to obtain this data had dynamic filters that switched the year of the displayed data. Due to a lack fo static data, we manually scraped the tables for previous years and saved the data in a CSV format

```{r importing csv}
# 2018 Data
revenue_growth_18 = read_csv("./gambling_revenue_data/2018_monthly_revenue.csv")
state_revenue_18  = read_csv("./gambling_revenue_data/2018_state_revenue.csv")

# 2019 Data
revenue_growth_19 = read_csv("./gambling_revenue_data/2019_monthly_revenue.csv")
state_revenue_19  = read_csv("./gambling_revenue_data/2019_state_revenue.csv")

# 2020 Data
revenue_growth_20 = read_csv("./gambling_revenue_data/2020_monthly_revenue.csv")
state_revenue_20  = read_csv("./gambling_revenue_data/2020_state_revenue.csv")

# 2021 Data
revenue_growth_21 = read_csv("./gambling_revenue_data/2021_monthly_revenue.csv")
state_revenue_21  = read_csv("./gambling_revenue_data/2021_state_revenue.csv")

# 2022 Data
revenue_growth_22 = read_csv("./gambling_revenue_data/2022_monthly_revenue.csv")
state_revenue_22  = read_csv("./gambling_revenue_data/2022_state_revenue.csv")

# 2023 Data
revenue_growth_23 = read_csv("./gambling_revenue_data/2023_monthly_revenue.csv")
state_revenue_23  = read_csv("./gambling_revenue_data/2023_state_revenue.csv")

# 2024 Data
revenue_growth_24 = read_csv("./gambling_revenue_data/2024_monthly_revenue.csv")
```
### cleaning the url scraped data so that it matches data types with the manually scraped data

```{r cleaning}
# --- Clean Scraped Data to match CSV format ---
# Function to remove '$', ',' and '%', convert to numeric and to handle '-' corresponding to missing data
clean_money_col = function(x) {
  # Remove currency symbols
  x_clean = gsub("[$,%]", "", x)
  # Replace dashes or empty strings with NA explicitly
  x_clean[x_clean == "-" | x_clean == ""] = NA
  as.numeric(x_clean)
}

# Clean the Scraped 2024 State Data to match CSV data types
state_revenue_24_clean = state_revenue_24 |> 
  mutate(
    across(c(Handle, Revenue, Taxes, Hold), clean_money_col), # Convert money columns
  )

# Clean the Scraped 2025 Monthly Data to match CSV data types
revenue_growth_25_clean = revenue_growth_25 |> 
  mutate(
    across(c(Handle, Revenue, Taxes, Hold), clean_money_col)
  )
```

```{r merging}
# --- Create Master State Revenue File ---
total_state_revenue_raw = bind_rows(
  state_revenue_18 |> mutate(Year = 2018, Hold = Hold * 100),
  state_revenue_19 |> mutate(Year = 2019, Hold = Hold * 100),
  state_revenue_20 |> mutate(Year = 2020, Hold = Hold * 100),
  state_revenue_21 |> mutate(Year = 2021, Hold = Hold * 100),
  state_revenue_22 |> mutate(Year = 2022, Hold = Hold * 100),
  state_revenue_23 |> mutate(Year = 2023, Hold = Hold * 100),
  state_revenue_24_clean |> mutate(Year = 2024)
) |> 
  relocate(Year, .before = State) # Move Year to the front

# Filter out individual yearly "Total" rows
total_state_revenue = total_state_revenue_raw |> 
  filter(State != "Total") |> 
  relocate(Year, .before = State)

# Calculate the True Grand Total
state_grand_total = total_state_revenue |> 
  summarize(
    Year = NA, # Grand total spans all years
    State = "Total",
    Handle = sum(Handle, na.rm = TRUE),
    Revenue = sum(Revenue, na.rm = TRUE),
    Taxes = sum(Taxes, na.rm = TRUE),
    Hold = sum(Revenue, na.rm = TRUE) / sum(Handle, na.rm = TRUE)
  )

# Bind the Grand Total to the clean data
final_state_revenue = bind_rows(total_state_revenue, state_grand_total)

# --- Create Master Monthly Revenue File ---
total_revenue_growth_raw = bind_rows(
  revenue_growth_18 |> mutate(Hold = Hold * 100),
  revenue_growth_19 |> mutate(Hold = Hold * 100),
  revenue_growth_20 |> mutate(Hold = Hold * 100),
  revenue_growth_21 |> mutate(Hold = Hold * 100),
  revenue_growth_22 |> mutate(Hold = Hold * 100),
  revenue_growth_23 |> mutate(Hold = Hold * 100),
  revenue_growth_24 |> mutate(Hold = Hold * 100),
  revenue_growth_25_clean
)

# Filter out individual yearly "Total" rows
total_revenue_growth = total_revenue_growth_raw |> 
  filter(Month != "Total")

# Calculate the True Grand Total
monthly_grand_total = total_revenue_growth |> 
  summarize(
    Month = "Total",
    Handle = sum(Handle, na.rm = TRUE),
    Revenue = sum(Revenue, na.rm = TRUE),
    Taxes = sum(Taxes, na.rm = TRUE),
    Hold = (sum(Revenue, na.rm = TRUE) / sum(Handle, na.rm = TRUE)) * 100
  )

# Bind the Grand Total to the clean data
final_revenue_growth = bind_rows(total_revenue_growth, monthly_grand_total)
```

## Census Info

```{r}
library(tidycensus)
library(janitor)
```

```{r Income and Population}
vars_base = c(
  median_income = "B19013_001",
  total_pop = "B01003_001"
)

acs_base_2024 =
  get_acs(
    geography = "state",
    variables = vars_base,
    year = 2024,
    survey = "acs1"
  ) |>
  select(NAME, variable, estimate) |>
  pivot_wider(names_from = variable, values_from = estimate) |>
  clean_names()
```


```{r Education (% with bachelor’s degree or higher)}
# Pull raw ACS education data
educ_2024 = get_acs(
  geography = "state",
  variables = c(
    total = "B15003_001",    # total population age 25+
    ba1 = "B15003_022",      # BA degree
    ba2 = "B15003_023",      # Master's
    ba3 = "B15003_024",      # Professional degree
    ba4 = "B15003_025"       # Doctorate
  ),
  survey = "acs1",
  year = 2024
)

# Clean + compute % bachelor’s degree or higher
educ_clean_2024 =
  educ_2024 |>
  group_by(NAME) |>
  summarize(
    total = estimate[variable == "total"],
    ba_plus = sum(estimate[variable %in% c("ba1", "ba2", "ba3", "ba4")]),
    .groups = "drop"
  ) |>
  mutate(pct_bachelors = ba_plus / total)

```

```{r Age}
age_2024 = get_acs(
  geography = "state",
  variables = c(
    total = "B01001_001",
    male_20_34 = paste0("B01001_", sprintf("%03d", 8:10)),
    female_20_34 = paste0("B01001_", sprintf("%03d", 32:34))
  ),
  survey = "acs1",
  year = 2024
)

age_clean_2024 =
  age_2024 |>
  mutate(
    group = case_when(
      variable == "total" ~ "total",
      str_starts(variable, "male_20_34") ~ "young",
      str_starts(variable, "female_20_34") ~ "young",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(group)) |>
  group_by(NAME, group) |>
  summarize(estimate = sum(estimate), .groups = "drop") |>
  pivot_wider(names_from = group, values_from = estimate) |>
  mutate(pct_age_20_34 = young / total)

median_age_2024 = get_acs(
  geography = "state",
  variables = c(median_age = "B01002_001"),
  survey = "acs1",
  year = 2024
) |>
  select(NAME, median_age = estimate)

```

```{r Urbanization (% living in urban areas)}
urban_raw <-read_csv("Census_Urbanization_Data/DECENNIALDHC2020.P2-Data.csv")

urban_clean_2020 =
  urban_raw |>
  filter(str_detect(GEO_ID, "^040")) |>                      # keep States only
  transmute(
    State = NAME,
    total = as.numeric(P2_001N),
    urban = as.numeric(P2_002N),
    rural = as.numeric(P2_003N),
    pct_urban = urban / total,
    pct_rural = rural / total
  )
#need this double checked
```

```{r Poverty (% of population below poverty line)}
poverty_2024 =
  get_acs(
    geography = "state",
    variables = c(total = "B17001_001",
                  below_pov = "B17001_002"),
    survey = "acs1",
    year = 2024
  )

poverty_clean_2024 =
  poverty_2024 |>
  group_by(NAME) |>
  summarize(
    total = sum(estimate[variable == "total"]),
    poverty = sum(estimate[variable == "below_pov"]),
    .groups = "drop"
  ) |>
  mutate(
    pct_poverty = poverty / total
  )
```

```{r % Male}
# % Male
sex_2024 = get_acs(
  geography = "state",
  variables = c(
    male = "B01001_002",
    total = "B01001_001"
  ),
  survey = "acs1",
  year = 2024
)

sex_clean_2024 =
  sex_2024 |>
  group_by(NAME) |>
  summarize(
    male = sum(estimate[variable == "male"]),
    total = sum(estimate[variable == "total"]),
    .groups = "drop"
  ) |>
  mutate(pct_male = male / total)

```

```{r Unemployment Rate}
# Unemployment rate
unemp_2024 = get_acs(
  geography = "state",
  variables = c(
    labor_force = "B23025_003",   # In labor force
    unemployed = "B23025_005"     # Unemployed
  ),
  survey = "acs1",
  year = 2024
)

unemp_clean_2024 =
  unemp_2024 |>
  group_by(NAME) |>
  summarize(
    labor_force = sum(estimate[variable == "labor_force"]),
    unemployed = sum(estimate[variable == "unemployed"]),
    .groups = "drop"
  ) |>
  mutate(unemployment_rate = unemployed / labor_force)

```

```{r Race}
race_2024 = get_acs(
  geography = "state",
  variables = c(
    total = "B02001_001",
    white = "B02001_002",
    black = "B02001_003",
    asian = "B02001_005",
    hispanic = "B03003_003"
  ),
  survey = "acs1",
  year = 2024
)

race_clean_2024 =
  race_2024 |>
  group_by(NAME) |>
  summarize(
    total = sum(estimate[variable == "total"]),
    white = sum(estimate[variable == "white"]),
    black = sum(estimate[variable == "black"]),
    asian = sum(estimate[variable == "asian"]),
    hispanic = sum(estimate[variable == "hispanic"]),
    .groups = "drop"
  ) |>
  mutate(
    pct_white = white / total,
    pct_black = black / total,
    pct_asian = asian / total,
    pct_hispanic = hispanic / total
  )

```

```{r Internet Access}
internet_2024 = get_acs(
  geography = "state",
  variables = c(
    total = "B28002_001",
    broadband = "B28002_004"
  ),
  survey = "acs1",
  year = 2024
)

internet_clean_2024 =
  internet_2024 |>
  group_by(NAME) |>
  summarize(
    total = sum(estimate[variable == "total"]),
    broadband = sum(estimate[variable == "broadband"]),
    .groups = "drop"
  ) |>
  mutate(pct_broadband = broadband / total)

```

```{r GINI Index}
# here is the def if you do not know what this is: The Gini index, or Gini coefficient, is a measure of income, wealth, or consumption inequality within a population, ranging from 0 (perfect equality) to 1 (perfect inequality)

gini_2024 = get_acs(
  geography = "state",
  variables = c(gini = "B19083_001"),
  survey = "acs1",
  year = 2024
) |>
  select(NAME, gini = estimate)

```

```{r Renting}
rent_2024 = get_acs(
  geography = "state",
  variables = c(
    total = "B25003_001",
    owner = "B25003_002",
    renter = "B25003_003"
  ),
  survey = "acs1",
  year = 2024
)

rent_clean_2024 =
  rent_2024 |>
  group_by(NAME) |>
  summarize(
    total = sum(estimate[variable == "total"]),
    owner = sum(estimate[variable == "owner"]),
    renter = sum(estimate[variable == "renter"]),
    .groups = "drop"
  ) |>
  mutate(
    pct_rent = renter / total,
    pct_own = owner / total
  )
```


```{r Merge All Demographics + Gambling Revenue Together}
#Rename so all are State
acs_base_2024 = acs_base_2024 |> rename(State = name)
educ_clean_2024 = educ_clean_2024 |> rename(State = NAME)
age_clean_2024 = age_clean_2024 |> rename(State = NAME)
poverty_clean_2024 = poverty_clean_2024 |> rename(State = NAME)
sex_clean_2024 = sex_clean_2024 |> rename(State = NAME)
unemp_clean_2024 = unemp_clean_2024 |> rename(State = NAME)
race_clean_2024 = race_clean_2024 |> rename(State = NAME)
median_age_2024 = median_age_2024 |> rename(State = NAME)
internet_clean_2024 = internet_clean_2024 |> rename(State = NAME)
gini_2024 = gini_2024 |> rename(State = NAME)
rent_clean_2024 = rent_clean_2024 |> rename(State = NAME)

# Final combined demographic dataset
demographics_2024 =
  acs_base_2024 |> 
  left_join(educ_clean_2024 |> 
              select(State, pct_bachelors), 
            by = "State") |>
  left_join(age_clean_2024 |> 
              select(State, pct_age_20_34), 
            by = "State") |>
  left_join(urban_clean_2020 |> 
              select(State, pct_urban, pct_rural), 
            by = "State") |>
  left_join(poverty_clean_2024 |> 
              select(State, pct_poverty), 
            by = "State") |>
  left_join(sex_clean_2024 |> 
              select(State, pct_male), 
            by = "State") |>
  left_join(unemp_clean_2024 |> 
              select(State, unemployment_rate), 
            by = "State") |>
  left_join(race_clean_2024 |> 
              select(State, pct_white, pct_black, pct_asian, pct_hispanic), 
            by = "State") |>
  left_join(median_age_2024 |> 
              select(State, median_age), 
            by = "State") |>
  left_join(internet_clean_2024 |> 
              select(State, pct_broadband), 
            by = "State") |>
  left_join(gini_2024 |> 
              select(State, gini), 
            by = "State") |>
  left_join(rent_clean_2024 |> 
              select(State, pct_rent, pct_own), 
            by = "State")

# Merge with gambling revenue
revenue_2024 =
  final_state_revenue |>
  filter(Year == 2024, State != "Total") |>
  select(State, Revenue, Handle, Taxes, Hold)

model_data_2024 =
  revenue_2024 |>
  left_join(demographics_2024, by = "State")
```

```{r Look at Final Dataset}
#view(demographics_2024)
summary(demographics_2024)
glimpse(demographics_2024)

#view(model_data_2024)
summary(model_data_2024)
glimpse(model_data_2024)
```

```{r Clean Final Dataset}
model_clean_2024 <-
  model_data_2024 |>
  drop_na(Revenue, Handle, Taxes, Hold)

#See final Amount
nrow(model_clean_2024)
unique(model_clean_2024$State)

#See who was dropped
anti_join(model_data_2024, model_clean_2024, by = "State") |>
  select(State)
```

## Building predictive model

```{r}

# Build training dataset
training_data <- model_clean_2024 |>
  select(
    State,
    Revenue,
    total_pop,
    median_income,
    pct_bachelors,
    pct_age_20_34,
    pct_urban,
    pct_poverty
  ) |>
  drop_na()   # make sure no missing values in predictors or Revenue

```

```{r}

# Fit linear regression model

# Model: Revenue = β0 + β1 * total_pop + β2 * median_income + ... + error
fit_lm <- lm(
  Revenue ~ total_pop + median_income + pct_bachelors +
    pct_age_20_34 + pct_urban + pct_poverty,
  data = training_data
)

summary(fit_lm)

```

```{r}
library(dplyr)

states_no_gambling <-
  demographics_2024 |>
  anti_join(training_data, by = "State")

newdata_illegal <-
  states_no_gambling |>
  select(
    State,
    total_pop,
    median_income,
    pct_bachelors,
    pct_age_20_34,
    pct_urban,
    pct_poverty
  ) |>
  drop_na()  # just in case a state is missing some demographic

```

```{r}

# Predict revenue if these states legalized gambling

pred_illegal <-
  newdata_illegal |>
  mutate(
    predicted_revenue = predict(fit_lm, newdata = newdata_illegal)
  ) |>
  arrange(desc(predicted_revenue))

#view(pred_illegal)

```

```{r}
# Compare distribution of real vs predicted
summary(training_data$Revenue)
summary(pred_illegal$predicted_revenue)

```

```{r}

# Fit LASSO regression model

# LASSO requires glmnet + matrix inputs
library(glmnet)

# Build design matrix (same predictors as your LM model)
X_train <- model.matrix(
  Revenue ~ total_pop + median_income + pct_bachelors +
    pct_age_20_34 + pct_urban + pct_poverty,
  data = training_data
)[, -1]   # remove intercept column (glmnet adds its own)

# Outcome vector
y_train <- training_data$Revenue

set.seed(123)

# Cross-validated LASSO to choose lambda
lasso_cv <- cv.glmnet(
  X_train,
  y_train,
  alpha = 1,          # LASSO penalty
  nfolds = 5,
  standardize = TRUE  # recommended because predictors differ in scale
)

# Final LASSO model at best lambda
lasso_model <- glmnet(
  X_train,
  y_train,
  alpha = 1,
  lambda = lasso_cv$lambda.min,
  standardize = TRUE
)

lasso_cv$lambda.min #This is the penalty 

```

```{r Predictive model}
# Build the prediction matrix for the non-gambling states
X_new <- model.matrix(
  ~ total_pop + median_income + pct_bachelors +
    pct_age_20_34 + pct_urban + pct_poverty,
  data = newdata_illegal
)[, -1]   # remove intercept column

# LASSO predictions
pred_illegal_lasso <-
  newdata_illegal |>
  mutate(
    predicted_revenue_lasso = as.numeric(
      predict(lasso_model, newx = X_new)
    )
  ) |>
  arrange(desc(predicted_revenue_lasso))

pred_illegal_lasso <- pred_illegal_lasso |> 
  mutate(predicted_revenue_lasso = pmax(predicted_revenue_lasso, 0))


#View(pred_illegal_lasso)

```

```{r Compare models}
summary(training_data$Revenue)
summary(pred_illegal$predicted_revenue)          # from LM model
summary(pred_illegal_lasso$predicted_revenue_lasso)  # from LASSO model

#LASSO better
```

## Some tables and Figures

```{r Tables}
library(scales)
library(knitr)
library(kableExtra)
library(dplyr)

# Clean negative values
pred_illegal_lasso <- pred_illegal_lasso |>
  mutate(predicted_revenue_lasso = pmax(predicted_revenue_lasso, 0))

# Clean and format final table
pred_table <- pred_illegal_lasso |>
  mutate(
    Predicted_Revenue = dollar(predicted_revenue_lasso),
    Population = comma(total_pop),
    Median_Income = dollar(median_income)
  ) |>
  select(
    State,
    Population,
    Median_Income,
    pct_bachelors,
    pct_age_20_34,
    pct_urban,
    pct_poverty,
    Predicted_Revenue
  )

kable(pred_table, format = "html", caption = "Predicted Annual Gambling Revenue for Non-Legal States (LASSO Model)") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

```

```{r Bar Plot}
library(ggplot2)

pred_illegal_lasso_plot <- pred_illegal_lasso |>
  mutate(predicted_revenue_lasso = pmax(predicted_revenue_lasso, 0)) |>
  arrange(predicted_revenue_lasso) |>
  mutate(State = factor(State, levels = State))

ggplot(pred_illegal_lasso_plot, aes(x = State, y = predicted_revenue_lasso)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Predicted Annual Gambling Revenue (if legalized)",
    subtitle = "LASSO model predictions for states without legal sports gambling",
    x = "",
    y = "Predicted Revenue"
  ) +
  theme_minimal(base_size = 13)

```

```{r}
compare_df <- tibble(
  group = c(rep("Actual Revenue (Legal States)", nrow(training_data)),
            rep("Predicted Revenue (Non-Legal States)", nrow(pred_illegal_lasso))),
  revenue = c(training_data$Revenue,
              pred_illegal_lasso$predicted_revenue_lasso)
)

ggplot(compare_df, aes(x = revenue, fill = group)) +
  geom_density(alpha = 0.4) +
  scale_x_continuous(labels = dollar) +
  labs(
    title = "Distribution of Actual vs Predicted Gambling Revenue",
    x = "Revenue",
    y = "Density",
    fill = ""
  ) +
  theme_minimal(base_size = 13)

```

```{r}
scatter_df <- bind_rows(
  training_data |> mutate(type = "Actual Revenue"),
  pred_illegal_lasso |> 
    mutate(Revenue = predicted_revenue_lasso,
           type = "Predicted Revenue")
)

ggplot(scatter_df, aes(x = total_pop, y = Revenue, color = type)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Population vs Gambling Revenue",
    subtitle = "Comparing actual and predicted states",
    x = "Total Population",
    y = "Revenue",
    color = ""
  ) +
  theme_minimal(base_size = 13)

```

```{r}
summary_table <- tibble(
  Metric = c("Min", "1st Quartile", "Median", "Mean", "3rd Quartile", "Max"),
  Actual_Revenue = summary(training_data$Revenue)[c(1,2,3,4,5,6)],
  Predicted_Revenue = summary(pred_illegal_lasso$predicted_revenue_lasso)[c(1,2,3,4,5,6)]
)

kable(summary_table, format = "html", caption = "Distribution Comparison: Actual vs Predicted Revenue") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

```

```{r}
library(dplyr)

# 1. Prepare actual revenue table
actual_table <- training_data %>%
  select(State, Revenue) %>%
  mutate(
    revenue_source = "Actual (Legal States)",
    Revenue = as.numeric(Revenue)
  )

# 2. Prepare predicted revenue table (LASSO)
predicted_table <- pred_illegal_lasso %>%
  select(State, predicted_revenue_lasso) %>%
  rename(Revenue = predicted_revenue_lasso) %>%
  mutate(
    revenue_source = "Predicted (Non-Legal States)"
  )

# 3. Combine both tables
ranked_table <- bind_rows(actual_table, predicted_table) %>%
  arrange(desc(Revenue)) %>%
  mutate(
    Rank = row_number(),
    Revenue = scales::dollar(Revenue)
  ) %>%
  select(Rank, State, Revenue, revenue_source)

#view(ranked_table)

```

