---
title: "All In! Forecasting Sports Gambling Revenue in Illegal Markets"
---

```{r}
library(tidyverse)
library(leaflet)
library(sf)
library(maps)
library(crosstalk)

# Set default figure options
knitr::opts_chunk$set(
  fig.width = 6,
  out.width = "90%"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Motivation

Scandals surrounding gambling in sports have reached the national stage in recent months, fueled by a meteoric rise over the last few years that has seemingly gone unchecked by our regulators. Whether you watch sports, TV, streaming, use social media, or take the subway, you've definitely been exposed to sports betting advertisements in one way or another. Our analysis focuses on the demographic factors underlying revenue earned in legal sports betting markets, including education & income status, age, and proximity to a major sports team. We then built a model to project the revenue generated in states with illegal markets and those likely to legalize soon—such as California. The overarching goal is to illustrate how the potential profitability of the sports gambling industry might compromise regulatory oversight. 

Here's a high-level overview of our project: 
<iframe width="560" height="315" src="XYZ" frameborder="0" allowfullscreen></iframe>

## Map 

```{r}
# Turn off spherical geometry processing.
sf_use_s2(FALSE)

# Import states' revenue data 
total_state_revenue <- read_csv("data/total_state_revenue.csv")

# Get US State boundaries (Shapes)
# Set the CRS to 4326 (Lat/Lon) to ensure Leaflet renders it correctly
us_states_sf <- st_as_sf(map("state", plot = FALSE, fill = TRUE)) %>%
  st_set_crs(4326) %>%
  mutate(ID = str_remove(ID, ":.*"))

# 2. Prepare Data
# Create a match_name column to join with the state shapes
revenue_map_data <- total_state_revenue %>%
  mutate(match_name = tolower(State))

# Join Shapes with Data
# Duplicates the state shape for EVERY year of data available.
map_data <- us_states_sf %>%
  left_join(revenue_map_data, by = c("ID" = "match_name")) %>%
  filter(!is.na(Year)) %>% # Ensure we don't have empty years
  # Create a unique ID for every single row (State + Year) for crosstalk, separating the layers so the slider can toggle them
  mutate(unique_key = paste(State, Year))

# Create "Shared Data"
# Set the 'key' so crosstalk knows how to toggle rows uniqueely
shared_map_data <- SharedData$new(map_data, key = ~unique_key)

# Color Palette
pal <- colorNumeric(palette = "viridis", domain = map_data$Revenue, na.color = "white")

# Render the Interactive Map with Slider
bscols(
  widths = c(12, 12),
  
  # The Slider
  filter_slider("year", "Select Year", shared_map_data, ~Year, step = 1, round = TRUE, animate = TRUE),
  
  # The Map
  leaflet(shared_map_data) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    setView(lng = -96, lat = 37.8, zoom = 4) %>%
    addPolygons(
      fillColor = ~pal(Revenue),
      weight = 2,              # Thicker border
      opacity = 1,
      color = "black",         # Black border for high contrast
      dashArray = "",          # Solid line (removed dash)
      fillOpacity = 0.7,
      highlightOptions = highlightOptions(
        weight = 3,
        color = "white",       # Highlight border becomes white on hover
        dashArray = "",
        fillOpacity = 0.9,
        bringToFront = TRUE
      ),
      label = ~paste0(
        "<strong>", str_to_title(ID), "</strong> (", Year, ")<br/>",
        "Revenue: $", format(Revenue, big.mark = ",", scientific = FALSE), "<br/>",
        "Taxes: $", format(Taxes, big.mark = ",", scientific = FALSE)
      ) %>% lapply(htmltools::HTML),
      labelOptions = labelOptions(
        style = list("font-weight" = "normal", padding = "3px 8px"),
        textsize = "15px",
        direction = "auto"
      )
    ) %>%
    addLegend(
      pal = pal, 
      values = map_data$Revenue, 
      opacity = 0.7, 
      title = "Revenue",
      position = "bottomright"
    )
)
```



