---
title: "gambling_rev"
output: github_document
date: "2025-11-30"
---
```{r packages}
library(tidyverse)
library(rvest)
library(httr)
library(jsonlite)

knitr::opts_chunk$set(
fig.width = 6,
fig.asp = .6,
out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
ggplot2.continuous.colour = "viridis",
ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```
## Data scraping gambling revenue and growth
```{r scraping}
url = "https://rg.org/statistics/us"
gambling_html = read_html(url)
```

```{r}
gambling_html |>
  html_table()
```
### pull in 2024 revenue by state table 
```{r state revenue}
state_revenue_24 = 
  gambling_html |> 
  html_table() |> 
  first() 
```
### pull in 2025 revenue growth by month
```{r monthly growth}
revenue_growth_25= 
  gambling_html |> 
  html_table() |> 
  purrr::pluck(2)
```

The website that was scrapped to obtain this data had dynamic filters that switched the year of the displayed data. Due to a lack fo static data, we manually scraped the tables for previous years and saved the data in a CSV format

```{r importing csv}
# 2018 Data
revenue_growth_18 = read_csv("./gambling_revenue_data/2018_monthly_revenue.csv")
state_revenue_18  = read_csv("./gambling_revenue_data/2018_state_revenue.csv")

# 2019 Data
revenue_growth_19 = read_csv("./gambling_revenue_data/2019_monthly_revenue.csv")
state_revenue_19  = read_csv("./gambling_revenue_data/2019_state_revenue.csv")

# 2020 Data
revenue_growth_20 = read_csv("./gambling_revenue_data/2020_monthly_revenue.csv")
state_revenue_20  = read_csv("./gambling_revenue_data/2020_state_revenue.csv")

# 2021 Data
revenue_growth_21 = read_csv("./gambling_revenue_data/2021_monthly_revenue.csv")
state_revenue_21  = read_csv("./gambling_revenue_data/2021_state_revenue.csv")

# 2022 Data
revenue_growth_22 = read_csv("./gambling_revenue_data/2022_monthly_revenue.csv")
state_revenue_22  = read_csv("./gambling_revenue_data/2022_state_revenue.csv")

# 2023 Data
revenue_growth_23 = read_csv("./gambling_revenue_data/2023_monthly_revenue.csv")
state_revenue_23  = read_csv("./gambling_revenue_data/2023_state_revenue.csv")

# 2024 Data
revenue_growth_24 = read_csv("./gambling_revenue_data/2024_monthly_revenue.csv")
```
### cleaning the url scraped data so that it matches data types with the manually scraped data

```{r cleaning}
# --- Clean Scraped Data to match CSV format ---
# Function to remove '$', ',' and '%', convert to numeric and to handle '-' corresponding to missing data
clean_money_col = function(x) {
  # Remove currency symbols
  x_clean = gsub("[$,%]", "", x)
  # Replace dashes or empty strings with NA explicitly
  x_clean[x_clean == "-" | x_clean == ""] = NA
  as.numeric(x_clean)
}

# Clean the Scraped 2024 State Data to match CSV data types
state_revenue_24_clean = state_revenue_24 |> 
  mutate(
    across(c(Handle, Revenue, Taxes, Hold), clean_money_col), # Convert money columns
  )

# Clean the Scraped 2025 Monthly Data to match CSV data types
revenue_growth_25_clean = revenue_growth_25 |> 
  mutate(
    across(c(Handle, Revenue, Taxes, Hold), clean_money_col)
  )
```

```{r merging}
# --- Create Master State Revenue File ---
total_state_revenue_raw = bind_rows(
  state_revenue_18 |> mutate(Year = 2018, Hold = Hold * 100),
  state_revenue_19 |> mutate(Year = 2019, Hold = Hold * 100),
  state_revenue_20 |> mutate(Year = 2020, Hold = Hold * 100),
  state_revenue_21 |> mutate(Year = 2021, Hold = Hold * 100),
  state_revenue_22 |> mutate(Year = 2022, Hold = Hold * 100),
  state_revenue_23 |> mutate(Year = 2023, Hold = Hold * 100),
  state_revenue_24_clean |> mutate(Year = 2024)
) |> 
  relocate(Year, .before = State) # Move Year to the front

# Filter out individual yearly "Total" rows
total_state_revenue = total_state_revenue_raw |> 
  filter(State != "Total") |> 
  relocate(Year, .before = State)

# Calculate the True Grand Total
state_grand_total = total_state_revenue |> 
  summarize(
    Year = NA, # Grand total spans all years
    State = "Total",
    Handle = sum(Handle, na.rm = TRUE),
    Revenue = sum(Revenue, na.rm = TRUE),
    Taxes = sum(Taxes, na.rm = TRUE),
    Hold = sum(Revenue, na.rm = TRUE) / sum(Handle, na.rm = TRUE)
  )

# Bind the Grand Total to the clean data
final_state_revenue = bind_rows(total_state_revenue, state_grand_total)

# --- Create Master Monthly Revenue File ---
total_revenue_growth_raw = bind_rows(
  revenue_growth_18 |> mutate(Hold = Hold * 100),
  revenue_growth_19 |> mutate(Hold = Hold * 100),
  revenue_growth_20 |> mutate(Hold = Hold * 100),
  revenue_growth_21 |> mutate(Hold = Hold * 100),
  revenue_growth_22 |> mutate(Hold = Hold * 100),
  revenue_growth_23 |> mutate(Hold = Hold * 100),
  revenue_growth_24 |> mutate(Hold = Hold * 100),
  revenue_growth_25_clean
)

# Filter out individual yearly "Total" rows
total_revenue_growth = total_revenue_growth_raw |> 
  filter(Month != "Total")

# Calculate the True Grand Total
monthly_grand_total = total_revenue_growth |> 
  summarize(
    Month = "Total",
    Handle = sum(Handle, na.rm = TRUE),
    Revenue = sum(Revenue, na.rm = TRUE),
    Taxes = sum(Taxes, na.rm = TRUE),
    Hold = (sum(Revenue, na.rm = TRUE) / sum(Handle, na.rm = TRUE)) * 100
  )

# Bind the Grand Total to the clean data
final_revenue_growth = bind_rows(total_revenue_growth, monthly_grand_total)
```

## Some preliminary visuals

```{r Plotting gambling revenue by Year}
# Aggregate total revenue by year
yearly_revenue_summary = total_state_revenue |> 
  group_by(Year) |> 
  summarize(Total_Revenue = sum(Revenue, na.rm = TRUE))

yearly_revenue_summary |> 
  ggplot(aes(x = Year, y = Total_Revenue)) +
  geom_col(aes(fill = as.factor(Year)), show.legend = FALSE) +
  geom_text(aes(label = scales::dollar(Total_Revenue, scale = 1e-9, suffix = "B")), 
            vjust = -0.5) + # Add labels in Billions
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_x_continuous(breaks = 2018:2024) +
  labs(
    title = "Total US Gambling Revenue (2018-2024)",
    subtitle = "Aggregated revenue across all states",
    x = "Year",
    y = "Total Revenue"
  )
```


```{r Plotting gambling revenue by State Over time}
# Identify top 5 states by 2024 revenue to keep things looking clean
top_states = total_state_revenue |> 
  filter(Year == 2019) |> 
  slice_max(Revenue, n = 5) |> 
  pull(State)

state_growth_plot_data = total_state_revenue |> 
  mutate(
    State_Group = ifelse(State %in% top_states, State, "All Other")
  ) |> 
  group_by(Year, State_Group) |> 
  summarize(Revenue = sum(Revenue, na.rm = TRUE), .groups = "drop") |> 
  mutate(State_Group = fct_reorder(State_Group, Revenue, .desc = TRUE))


ggplot(state_growth_plot_data, aes(x = Year, y = Revenue, fill = State_Group)) +
  geom_col() +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_x_continuous(breaks = 2018:2024) +
  labs(
    title = "US Gambling Revenue Growth by State",
    x = "Year",
    y = "Revenue",
    fill = "State"
  )
```


```{r plotting revenue rank: top 5 gambling markets by year since 2018}
# Calculate Ranks for every state, every year
ranked_data = total_state_revenue |> 
  group_by(Year) |> 
  mutate(Rank = min_rank(desc(Revenue))) |> 
  ungroup()

# Identify "States of Interest"
# We want to track any state that was Top 5 at any point
focus_states = ranked_data |> 
  filter(Rank <= 5) |>  # can change if we want less point to start and end points: filter((Year == 2018 & Rank <= 5) | (Year == 2024 & Rank <= 5)) |>
  pull(State) |> 
  unique()

# Filter the dataset to just those states
plot_data = ranked_data |> 
  filter(State %in% focus_states)

plot_data |> 
  ggplot(aes(x = Year, y = Rank, color = State, group = State)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) + 
  scale_y_reverse(breaks = 1:max(plot_data$Rank, na.rm = TRUE)) + 
  scale_x_continuous(breaks = 2018:2024) +
  geom_text(data = plot_data |> filter(Year == 2018), 
            aes(label = paste(Rank, State)), x = 2017.8, hjust = 1, size = 3) +
  geom_text(data = plot_data |> filter(Year == 2024), 
            aes(label = paste(Rank, State)), x = 2024.2, hjust = 0, size = 3) +
  labs(
    title = "Shift in Top Gambling Markets (2018 vs 2024)",
    subtitle = "Tracking the rank of states that were in the top 5 in either 2018 or 2024",
    y = "Revenue Rank",
    x = "Year"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",        
    panel.grid.minor = element_blank()
  ) +
  coord_cartesian(xlim = c(2017, 2025))
```

